---
title: "Introduction to Tidyverse (using Genomic data)"
description: |
 RMarkdown file to accompany the Tidyverse for Genomics workshop
author:
  - name: Janani Ravi
    url: https://jravilab.github.io
    affiliation: PDI, MMG, MSU | R-Ladies East Lansing
    affiliation_url: https://rladies-eastlansing.github.io
    orcid_id: 0000-0001-7443-925X
  - name: Arjun Krishnan
    url: https://thekrishnanlab.org
    affiliation: CMSE & BMB, MSU
    affiliation_url: https://cmse.msu.edu/
    orcid_id: 0000-0002-7980-4110
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: TRUE
    toc_depth: 3
  pdf_document:
    toc: yes
  html_notebook:
    theme: flatly
    toc: yes
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
editor_options:
  chunk_output_type: console
always_allow_html: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  # comment = "##",
  R.options = list(width = 60)
)
```

### About me
I am a computational biologist, an Asst. Professor in the departments of Pathobiology & Diagnostic Investigation and Microbiology & Molecular Genetics at Michigan State University.
In our group, we develop computational approaches to understand infectious disease biology.
Check out my [webpage](https://jravilab.github.io) for more info.
You can reach me [here](https://jravilab.github.io/#contact).

I'm also the founder & co-organizer of the [R-Ladies East Lansing](https://rladies-eastlansing.github.io), Women+ Data Science, and AsiaR.
We conduct R and data science related meetups & workshops regularly!
So, do check out our upcoming events on [Meetup](https://meetup.com/rladies-eastlansing/events).

# Part 1: Getting Started w/ `readr`
You can access all relevant material pertaining to this workshop [here](https://github.com/jananiravi/workshop-tidyverse).
Other related workshops & useful [cheatsheets](http://github.com/jananiravi/cheatsheets).

## Installation and set-up
### Install RStudio
*Running RStudio locally?*
[**Download RStudio**](http://rstudio.com/download)

*Want to try the latest 'Preview' version of RStudio?*
[**RStudio Preview version**]((https://www.rstudio.com/products/rstudio/download/preview/))

*Trouble with local installation?* Login & start using [**RStudio Cloud**](https://rstudio.cloud) right away!

*New to RStudio IDE?*
Use Help [Page #1](https://www.rstudio.com/products/rstudio/#Desktop) &
[Page #2](https://www.rstudio.com/products/rstudio/features/)


### Install R
*... if you haven't already!*
The RStudio startup message should specify your current local version of R.
For _e.g.,_ `R v4.1.2`

- [Download R](https://cran.r-project.org)
- [*Update to the latest version of R?*](https://www.linkedin.com/pulse/3-methods-update-r-rstudio-windows-mac-woratana-ngarmtrakulchol)
- [A Windows-specific solution to updating R](https://www.r-statistics.com/2015/06/a-step-by-step-screenshots-tutorial-for-upgrading-r-on-windows/)

### Install packages
- **Install**: once
- **Load**: each time

```{r installation, eval=F, echo=T}
install.packages("tidyverse")   # for data wrangling
install.packages("here")        # to set paths relative to your current project.
# install.packages("gapminder") # sample dataset
```
Explore the [RLEL workshops](https://github.com/rladies-eastlansing/meetup-presentations) for more examples and sample codes for Tidy Data and DataViz [_e.g.,_ >> `presentations/20181105-workshop-tidydata` uses the `gapminder` & `USArrests` datasets].


**Trouble installing tidyverse?**

- Check your R/RStudio versions
- For the time-being install the individual packages using `install.packages("PACKAGENAME")`
- Check out the whole `tidyverse` suite of packages [here](https://www.tidyverse.org/packages/)

```{r tidyverse, eval=FALSE, echo=TRUE}
# If tidyverse installation fails, install individual constituent packages this way...
install.packages("readr")    # Importing data files
install.packages("tidyr")    # Tidy Data
install.packages("dplyr")    # Data manipulation
install.packages("ggplot2")  # Data Visualization (w/ Grammar of Graphics)
install.packages("readxl")   # Importing excel files
```

### Loading packages
```{r loading, eval=T, echo=TRUE}
library(tidyverse)
# OR load the individual packages, if you have trouble installing/loading `tidyverse`
# library(readr)
# library(readxl)
# library(tidyr)
# library(dplyr)
# library(ggplot2)
library(here) # https://github.com/jennybc/here_here
# library(gapminder) # useful dataset for data wrangling, visualization
```

## Some useful cheatsheets
[Cheatsheets \@RStudio](https://www.rstudio.com/resources/cheatsheets/) | [Our cheatsheets repo](https://github.com/jananiravi/cheatsheets)

More resources towards the [end of the document](#refs).

## Data import

- `read_csv`, `write_csv`
- `read_tsv`, `write_tsv`
- `read_delim`, `write_delim`
- `here::here`

```{r data-import, eval=T, echo=TRUE}
here::here() # where you opened your RStudio session/Project

# Downloading sample genomics dataset from NCBI's FTP site
url <- "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE69nnn/GSE69360/suppl/GSE69360_RNAseq.counts.txt.gz"
gse69360 <- read_tsv(url) # to read tab-delimitted files
?read_tsv # to check defaults
```

```{r data-import2, eval=F, echo=T}
# Comma-separated values, as exported from excel/spreadsheets
read_csv(file="path/to/my_data.csv", col_names=T)
# Other atypical delimitters
read_delim(file="path/to/my_data.txt", col_names=T, delim="//")

# Other useful packages
# readxl by Jenny Bryan
read_excel(path="path/to/excel.xls",
           sheet=1,
           range="A1:D50",
           col_names=T)

## Tip: Always open .Rproj and use relative paths with here()
## Example with here()
read_tsv(file=here("genomics/GSE69360_RNAseq.counts.txt"), col_names=T)
```


## Knowing your data
*Dataset details*:

- AA: Agilent Adult; AF: Agilent Fetus
- BA: BioChain Adult; BF: BioChain Fetus
- OA: OriGene Adult
- Tissues: Colon, Heart, Kidney, Liver, Lung, Stomach

```{r data-structure, echo=TRUE, eval=F}
str(gse69360)              # Structure of the dataframe
gse69360                   # Data is in a cleaend up 'tibble' format by default
head(gse69360)             # Shows the top few observations (rows) of your data frame
glimpse(gse69360)          # Info-dense summary of the data
View(head(gse69360, 100))  # View data in a visual GUI-based spreadsheet-like format

colnames(gse69360)         # Column names
nrow(gse69360)             # No. of rows
ncol(gse69360)             # No. of columns

gse69360[1:5,7:10]         # Subsetting a dataframe

## saving the data file
write_tsv(gse69360[1:100,7:12], "gse_subset.txt")
```

```{r kable, layout="l-body-outset"}
library(knitr)
kable(head(gse69360))
```

```{r paged_table, layout="l-body-outset"}
library(rmarkdown)
paged_table(gse69360)
```

***
# Part 2: Reshaping data w/ `tidyr`
```{r tidyr, echo=TRUE, eval=FALSE}
## legacy tidyverse funcgions
gather()   # Gather COLUMNS -> ROWS
spread()   # Spread ROWS -> COLUMNS

## newer tidyverse functions
pivot_longer()  # Pivot data from wide to long
pivot_wider()   # Pivot data from long to wide

separate() # Separate 1 COLUMN -> many COLUMNS
unite()    # Unite several COLUMNS -> 1 COLUMN

```

## Gather, Spread
- `gather`: *Gather columns into key-value pairs. Wide -> Long*
- `spread`: *Spread a key-value pair across multiple columns: Long -> Wide*

```{r gather-spread, echo=T, eval=T}
# Gather all columns except 'Geneid'
gse69360 %>%
  select(Geneid, matches("[AF]_")) %>%
  # gather(-Geneid, key="Sample", value="Counts") # wide -> long format  ## legacy
  pivot_longer(cols=matches("[AF]_"), names_to="Sample", values_to="Counts")

# Gather, then Spread --> Back to original data
gse69360 %>%
  select(Geneid, matches("[AF]_")) %>%
  # gather(-Geneid, key="Sample", value="Counts") %>%
  # spread(key = "Sample", value = "Counts")     # spread to turn back to the original data!
  pivot_longer(cols=matches("[AF]_"), names_to="Sample", values_to="Counts") %>%
  pivot_wider(names_from="Sample", values_from="Counts")
```


## Unite, Separate
- `unite`:    *Unite multiple columns into one*
- `separate`: *Separate one column into multiple columns*

```{r separate-unite, echo=T, eval=T}
gse69360 %>%
  select(Geneid, matches("[AF]_")) %>%               # selecting only Counts columns
  pivot_longer(cols=matches("[AF]_"), names_to="Sample", values_to="Counts") %>% # wide -> long
  separate(Sample, into=c("Source_Stage", "Tissue"), sep="_") # separate logically

gse69360 %>%
  select(Geneid, matches("[AF]_")) %>%
  pivot_longer(cols=matches("[AF]_"), names_to="Sample", values_to="Counts") %>%
  separate(Sample, into=c("Source_Stage", "Tissue"), sep="_") %>%
  separate(Source_Stage, into=c("Source", "Stage"), sep=1) # separate by char position

gse69360 %>%
  select(Geneid, matches("[AF]_")) %>%
  pivot_longer(cols=matches("[AF]_"), names_to="Sample", values_to="Counts") %>%
  separate(Sample, into=c("Source_Stage", "Tissue"), sep="_") %>%
  separate(Source_Stage, into=c("Source", "Stage"), sep=1) %>%
  unite(Stage_Tissue, Stage, Tissue) # combining a different set of columns
```

***
# Part 3: Data wranging with `dplyr`
```{r dplyr, echo=TRUE, eval=FALSE}
conflicted::conflict_prefer(name="filter", winner="dplyr")
filter()    # PICK observations by their values | ROWS
select()    # PICK variables by their names | COLUMNS
mutate()    # CREATE new variables w/ functions of existing variables | COLUMNS
transmute() # COMPUTE 1 or more COLUMNS but drop original columns
arrange()   # REORDER the ROWS
summarize() # COLLAPSE many values to a single SUMMARY
group_by()  # GROUP data into rows with the same value of variable (COLUMN)
```

## Filter
- `filter`: *Return rows with matching conditions*

```{r filter, echo=TRUE, eval=TRUE}
head(gse69360) # Snapshot of the dataframe

# Now, filter by condition
filter(gse69360, Length<=50)

# Can be rewritten using "Piping" %>%
gse69360 %>%   # Pipe ('then') operator to serially connect operations
  filter(Length <= 50)

# Filtering using regex/substring match
gse69360 %>%
  filter(grepl("chrY", Chr))

# Two filters at a time
gse69360 %>%
  filter(Length <= 50 & grepl("chrY", Chr))

```

## Select
- `select`: *Select/rename variables/columns by name*

```{r select, echo=T, eval=T}
# Selecting columns that match a pattern
gse69360 %>%
  select(Geneid, matches(".F_"))

# Excluding specific columns
gse69360 %>%
  select(-Chr, -Start, -End, -Strand, -Length)

# Excluding columns matching a pattern
gse69360 %>%
  select(-matches("[AF]_"))

# Select then Filter
gse69360 %>%
  select(Geneid, Chr, Length, matches("[AF]_")) %>%
  filter(grepl("chrY", Chr) | Length <= 100)
```


## Mutate
- `mutate`: *Adds new variables; keeps existing variables*
- `transmute`: *Adds new variables; drops existing variables*

```{r mutate, echo=T, eval=T}
# Excluding columns matching a condition
gse69360 %>%
  select(-matches("[AF]_")) %>%
  head(., 10) #%>% View()

# Storing gene location information in a seprate data frame
gene_loc <- gse69360 %>%                                            # saving output to a variable
  select(-matches("[AF]_")) %>%                                     # select columns
  mutate(Geneid = gsub("\\.[0-9]*$", "", Geneid)) %>%               # remove isoform no.
  mutate(Chr = gsub(";.*$", "", gse69360$Chr)) %>%                  # keep the first element for Chr
  mutate(Start = as.numeric(gsub(";.*$", "", gse69360$Start))) %>%  # "" for Start
  mutate(End = as.numeric(gsub(";.*$", "", gse69360$End))) %>%      # "" for End
  mutate(Strand = gsub(";.*$", "", gse69360$Strand))                # "" for Strand

# Check to see if you have what you expected!
gene_loc
head(gene_loc, 10)

# Creating new variables
gene_loc %>%
  mutate(kbStart = Start/1000,     # creates new variables/columns
         kbEnd = End/1000,
         kbLength = Length/1000)

# Creating new variables & dropping old ones
gene_loc %>%
  transmute(kbStart = Start/1000,  # drops original columns
            kbEnd = End/1000,
            kbLength = Length/1000)

```

## Distinct & Arrange
- `distinct`: *Pick unique entries*
- `arrange`: *Arrange rows by variables*

```{r arrange, echo=TRUE, eval=TRUE}
# Pick only the unique entries in a column
gene_loc %>%
  distinct(Chr)

gene_loc %>%
  distinct(Strand)

# Pick unique combinations
gene_loc %>%
  distinct(Chr, Strand)

# Then sort aka arrange your data
gene_loc %>%
  arrange(desc(Chr))    # sort in descending order

gene_loc %>%
  arrange(Chr, Length)  # sort by Chr, then Length
# arrange(Chr, -Length) # to reverse sort by 'numeric' Length
```


## Group_by & Summarize
- `summarize`: *Reduces multiple values down to a single value*
- `group_by`: *Combine entries by one or more variables*

```{r summarize, eval=TRUE, echo=TRUE}
# Combine by a variable, then calculate summary statistics for each group
gene_loc %>%
  group_by(Chr) %>%                         # combine rows by Chr
  summarize(numGenes = n(),                 # then summarise, number of genes/Chr
            startoffirstGene = min(Start))  # min to get the first Start location

# Example to show you can use all math/stat functions to summarize data groups
gene_loc %>%
  arrange(Length) %>%
  group_by(Chr, Strand) %>%
  summarize(numGenes = n(),
            smallestGene = first(Geneid),
            minLength = min(Length),
            firstqLength = quantile(Length, 0.25),
            medianLength = median(Length),
            iqrLength = IQR(Length),
            thirdqLength = quantile(Length, 0.75),
            maxLength = max(Length),
            longestGene = last(Geneid))

# Renaming chromsosomes & declaring them as factors w/ an intrinsic order
# gene_loc$Chr <- factor(gene_loc$Chr,
#                        levels = paste("chr",
#                                       c((1:22), "X", "Y", "M"),
#                                       sep=""))

# Saving your data locally
gene_loc %>%
  write_tsv(here("genomics/GSE69360.gene-locations.txt"))

```

### More data wrangling
Let's combine everything from above to tidy the full GSE69360 dataset.

Dataset details:

- AA: Agilent Adult; AF: Agilent Fetus
- BA: BioChain Adult; BF: BioChain Fetus
- OA: OriGene Adult
- Tissues: Colon, Heart, Kidney, Liver, Lung, Stomach

```{r more-wrangling, echo=TRUE, eval=TRUE}
# Extracting just the expression values & cleaning it up
View(head(gse69360, 50))

gene_counts <- gse69360 %>%
  select(-Chr, -Start, -End, -Strand, -Length) %>%          # another way to select just the expression data
  rename(OA_Stomach = OA_Stomach1) %>%                      # rename couple of columns
  mutate(OA_Stomach2 = NULL, OA_Stomach3 = NULL) %>%        # remove a couple of columns
  mutate(Geneid = gsub("\\.[0-9]*$", "", Geneid))           # cleanup data a specific column

logcpm <- gene_counts %>%
  select(-Geneid) %>%
  mutate_all(function(x) { log2((x*(1e+6)/sum(x)) + 1) } )  # convert counts in each sample to counts-per-million

summary(logcpm)

gene_logcpm <- gene_counts %>%                              # a new dataframe with logcpm values
  select(Geneid) %>%
  bind_cols(logcpm) %>%                                     # bind the logcpm matrix to the geneids
  gather(-Geneid, key = "Sample", value = "Logcpm") %>%     # convert to tidy data
  separate(Sample,                                          # cleanup complex variables
           into = c("Source", "Stage", "Tissue"),
           sep = c(1,2),
           remove = F) %>%                                  # keep original variable
  mutate(Tissue = gsub("^_", "", Tissue),
         Stage = ifelse(Stage == "A", "Adult", "Fetus"))

View(head(gene_logcpm, 50))

# Plotting the distribution of gene-expression in each sample
gene_logcpm %>%
  ggplot(aes(x = Sample, y = Logcpm, color = Tissue, linetype = Stage)) +
  geom_boxplot(outlier.size = 0.2, outlier.shape = 0.2) +
  scale_y_continuous(limits = c(0, 1)) +
  coord_flip() +
  theme_minimal()
```

***
# Part 4: Visualizing tidy data w/ `ggplot`
## Basics of ggplot2
**Creating a plot w/ Grammar of Graphics**

> 1. Recap and continuation of **dplyr**
> 2. Basics of plotting data with **ggplot2**: `data`, `aes`, `geom`
> 3. Customization: Colors, labels, and legends


### Barplots & Histograms
- `ggplot`, `factor`, `aes`
- `geom_bar`, `geom_histogram`
- `facet_wrap`
- `scale_x_log10`, `labs`, `coord_flip`, `theme`, `theme_minimal`

```{r ggplot-bars-hist, echo=T, eval=T}
gene_loc %>%                                                      # data
  ggplot(aes(x = Chr)) +                                          # aesthetics: what to plot?
  geom_bar()                                                      # geometry: how to plot?

gene_loc$Chr <- factor(gene_loc$Chr,
                       levels = paste("chr",
                                      c((1:22), "X", "Y", "M"),
                                      sep=""))

plot_chr_numgenes <- gene_loc %>%
  ggplot(aes(x = Chr)) +
  geom_bar()
plot_chr_numgenes

plot_chr_numgenes +
  coord_flip() +
  theme_minimal()

plot_chr_numgenes +
  labs(title = "No. genes per chromosome",
       x = "Chromosome",
       y = "No. of genes") +
  theme_minimal() +
  coord_flip()

gene_loc %>%
  ggplot(aes(x = Length)) +
  geom_histogram(color = "white") +
  scale_x_log10() +
  theme_minimal()

plot_chr_genelength <- gene_loc %>%
  ggplot(aes(x = Length, fill = Chr)) +
  geom_histogram(color = "white") +
  scale_x_log10() +
  theme_minimal() +
  facet_wrap(~Chr, scales = "free_y")
plot_chr_genelength

plot_chr_genelength +
  theme(legend.position = "none") +
  labs(x = "Gene length (log-scale)",
       y = "No. of genes")

```


### Scatter plots
- `geom_point`
- `geom_abline`, `geom_vline`, `geom_hline`
- `geom_smooth`, `geom_text_repel`

```{r ggplot-scatter, echo=TRUE, eval=TRUE}

gene_loc %>%
  ggplot(aes(x = End-Start, y = Length)) +
  geom_point()

plot_strend_length <- gene_loc %>%
  ggplot(aes(x = End-Start, y = Length)) +
  geom_point(alpha = 0.1, size = 0.5, color = "grey", fill = "grey")
plot_strend_length

plot_strend_length <- plot_strend_length +
  scale_x_log10("End-Start") +
  scale_y_log10("Gene length") +
  theme_minimal()
plot_strend_length

plot_strend_length +
  geom_abline(intercept = 0, slope = 1, col = "red") +
  geom_hline(yintercept = 500, color = "blue") +
  geom_vline(xintercept = 1000, color = "orange")

gene_loc %>%
  group_by(Chr) %>%
  summarize(meanLength = mean(Length), numGenes = n())

gene_loc %>%
  group_by(Chr) %>%
  summarize(meanLength = mean(Length), numGenes = n()) %>%
  ggplot(aes(x = numGenes, y = meanLength)) +
  geom_point()

# install.packages("ggrepel", dependencies=T)
library(ggrepel)
gene_loc %>%
  group_by(Chr) %>%
  summarize(meanLength = mean(Length), numGenes = n()) %>%
  ggplot(aes(x = numGenes, y = meanLength)) +
  geom_point() +
  geom_smooth(color = "lightblue", alpha = 0.1) +
  labs(x = "No. of genes", y = "Mean gene length") +
  geom_text_repel(aes(label = Chr), color="red", segment.color="grey80") +
  theme_minimal()
```

### Boxplots & Violin plots
- `geom_boxplot`, `geom_violin`
- `scale_y_continuous`

```{r echo=TRUE, eval=TRUE}
gene_logcpm %>%
  ggplot(aes(x = Sample, y = Logcpm, color = Tissue, linetype = Stage)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()

gene_logcpm %>%
  ggplot(aes(x = Sample, y = Logcpm, color = Tissue, linetype = Stage)) +
  geom_violin() +
  scale_y_continuous(limits = c(0, 0.5)) +
  coord_flip() +
  theme_minimal()

# Plotting the distribution of gene-expression in each sample
plot_sample_bxp <- gene_logcpm %>%
  ggplot(aes(x = Sample, y = Logcpm, color = Tissue, linetype = Stage)) +
  geom_boxplot(outlier.size = 0.2, outlier.alpha = 0.2) +
  scale_y_continuous(limits = c(0, 1)) +
  coord_flip() +
  theme_minimal()
plot_sample_bxp

# Plotting scatterplot of 2 sets of samples
plot_ffcolon_scatter <- gene_logcpm %>%
  filter(Sample == "AF_Colon" | Sample == "BF_Colon") %>%
  select(Geneid, Sample, Logcpm) %>%
  spread(key = Sample, value = Logcpm) %>%
  ggplot(aes(x = AF_Colon, y = BF_Colon)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method=lm) +
  theme_minimal()
plot_ffcolon_scatter
```


### Some data sleuthing
```{r echo=TRUE, eval=TRUE}
# Finding genes with high variance across samples
num_totgenes <- gene_logcpm %>%
  distinct(Geneid) %>%
  nrow()

highvar_genes <- gene_logcpm %>%
  group_by(Geneid) %>%
  summarize(iqr = IQR(Logcpm)) %>%
  top_n((ceiling(num_totgenes*0.05)), iqr) %>%
  pull(Geneid)

length(highvar_genes)

# Plotting expression of high-var Y chr genes across samples
chry_highvar_genes <- gene_loc %>%
  filter(Chr == "chrY" & Geneid %in% highvar_genes) %>%
  pull(Geneid)

gene_logcpm %>%
  filter(Geneid %in% chry_highvar_genes)

plot_chry_highvar_boxplot <- gene_logcpm %>%
  filter(Geneid %in% chry_highvar_genes) %>%
  ggplot(aes(x = reorder(Sample, Logcpm, FUN = median),
             y = Logcpm,
             color = Sample)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none")
plot_chry_highvar_boxplot
```

***
# Part 5: Export & Wrap-up w/ `RMarkdown`
## Saving your plots
### ggsave
**Save a ggplot (or other grid object) with sensible defaults**

```{r ggsave, eval=T, echo=TRUE}
library(tidyverse)
# Save your file name
plot1 <- "chr_highvar.png"

# Save your absolute/relative path
my_full_path <- here("genomics")

# To save as a tab-delimited text file ...
ggsave(filename=plot1,
       plot=plot_chry_highvar_boxplot,
       device="png",
       path=my_full_path,
       dpi=600)

```

## Saving your data files
### write_delim
**Write a data frame to a delimited file**

```{r write_delim, eval=F, echo=TRUE}
library(tidyverse)
# Save your file name
filename <- "my_new_data.txt"

# Save your absolute/relative path
my_full_path <- here("data")

# To save as a tab-delimited text file ...
write_tsv(x=my_newly_formatted_data, # your final reformatted dataset
          path=paste(my_full_path, filename, "/"), # Absolute path recommended.
          # However, you can directly use 'filename' here
          # if you are saving the file in the same directory
          # as your code.
          col_names=T) # if you want the column names to be
# saved in the first row, recommended

# Alternatively, you could save it as a comma-separated text file
write_csv(x=my_newly_formatted_data,
          path=my_path,
          col_names=T)
# Or save it with any other delimiter
# choose wisely, pick a delim that's not part of your dataframe
write_delim(x=my_newly_formatted_data,
            path=my_path,
            col_names=T,
            delim="---")
```

## *What you learnt today!*

::: l-body-outset
| Option      | Description                                        |
|-------------|----------------------------------------------------|
|**Part 1** | **Getting Started**                                      |
| `install.packages`| Download and install packages from CRAN-like repositories or from local files |
| `library`   | Library and require load and attach add-on packages|
| `package::function`| To run a function w/o loading the package   |
|*Import* | `tidyverse` > `readr` & `readxl`                                     |
| `read_delim`| Read a delimited file (incl csv, tsv) into a tibble|
| `read_csv, read_tsv`  | read_csv() and read_tsv() are special cases of the general read_delim()|
| `read_excel`| Read xls and xlsx files                            |
| `here`| Part of `here` package. To set paths relative to your current project directory  |
|*Data snapshot*                                                |
| `str`       | Compactly Display the Structure of an Arbitrary R Object|
| `head`      | Return the First or Last Part of an Object         |
| `glimpse`   | Get a glimpse of your data                         |
| `View`      | Invoke a Data Viewer                               |
| `kable`     | Create tables in LaTeX, HTML, Markdown and reStructuredText|
| `paged_table`| Create a table in HTML with support for paging rows and columns|
|**Part 2** | `tidyverse` > `tidyr`                                   |
| `pivot_longer`    | Gather Columns Into Key-Value Pairs (COLS -> ROWS) |
| `pivot_wider`    | Spread a key-value pair across multiple columns    |
| `separate`  | Separate one column into multiple column           |
| `unite`     | Unite multiple columns into one                    |
|**Part 3** | `tidyverse` > `dplyr`                                       |
| `filter`    | Return rows with matching conditions               |
| `select`    | Select/rename variables by name                    |
| `mutate`    | Add new variables                                  |
| `transmute` | Add new variables & drops existing variables       |
| `distinct`  | Get unique entries                                 |
| `arrange`   | Arrange rows by variables                          |
| `summarise` | Reduces multiple values down to a single value     |
| `group_by`  | Group by one or more variables                     |
| `join`      | Join two tbls together: `left_join`, `right_join`, `inner_join`|
| `bind`      | Efficiently bind multiple data frames by row and column: `bind_rows`, `bind_cols`|
| `setops`    | Set operations: `intersect`, `union`, `setdiff`, `setequal`|
|**Part 4** | `tidyverse` > `ggplot`                                   |
| `ggplot`    | Create a new ggplot                                |
|*Aesthetics*|
| `aes`| Specify details on what to plot                            |
| `factor`| Specify a variable to be ordered & categorical          |
| `facet_wrap`| Split into multiple sub-plots based on a categorical variable |
|*Geometries*|
| `geom_bar`| Create a barplot                                |
| `geom_histogram`| Create a histogram                                |
| `geom_point`| Create a scatter plot                                |
| `geom_boxplot`| Create a boxplot                                |
| `geom_violin`| Create a violin plot                                |
| `geom_abline, geom_hline, geom_vline`| Add slanted, horizontal, or vertical lines |
| `geom_smooth`| Add a smooth curve that fits the data points                       |
|*Plot customization*|
| `theme, theme_minimal, theme_bw`| Adjust themes to minimal, black/white, etc |
| `scale_x_log10`| Change x/y axes to log scale                    |
| `scale_y_continuous`| Change x/y axes to continuous & set limits |
| `coord_flip`| Flip x & y coordinates                                |
| `labs`| Specify axes labels and plot titles                          |
|**Part 5** | **Export & Wrap-up**                                     |
|*Export* | `tidyverse` > `readr`                                            |
| `ggsave`    | Save a ggplot (or other grid object) with sensible defaults|
| `write_delim`| Write a data frame to a delimited file            |
| `write_tsv` | write_delim customized for tab-separated values    |
| `write_csv` | write_delim customized for comma-separated values  |
:::

***

## Credits
[Arjun Krishnan](https://thekrishnanlab.org) and [I](https://jravilab.github.io) co-developed the content for this workshop.

### Acknowledgements
- [Krishnan Lab](https://thekrishnanlab.org) | [JRaviLab](https://jravilab.github.io)
- [R-Ladies East Lansing](https://rladies-eastlansing.github.io)
- [R-Ladies Chennai](https://meetup.com/rladies-chennai) | [R-Ladies Bangalore](https://meetup.com/rladies-bangalore) | [R-Ladies Tunis](https://meetup.com/rladies-tunis)

### Contact
- Webpage: <https://jravilab.github.io> | <https://thekrishnanlab.org>
- Email: <janani@msu.edu> | <arjun@msu.edu>
- Twitter: [\@janani137](https://twitter.com/janani137) | [\@compbiologist](https://twitter.com/compbiologist)
- GitHub: [\@jananiravi](https://github.com/jananiravi) | [\@arjunkrish](https://github.com/arjunkrish)

## Additional resources {#refs}
- You can access all relevant material pertaining to this workshop [here](https://github.com/jananiravi/workshop-tidyverse).
- Other related [RLEL workshops](http://github.com/rladies-eastlansing/meetup-presentations) & useful [cheatsheets](http://github.com/jananiravi/cheatsheets).

### Some awesome open-source books
- Hands-On Programming with R: Grolemund #HOPR https://rstudio-education.github.io/hopr/
- R for Data Science: Wickham & Grolemund #R4DS https://r4ds.had.co.nz
- R Programming for Data Science: Peng https://leanpub.com/rprogramming
- Learning Statistics with R: Navarro https://learningstatisticswithr.com/book

### Other helpful links
- #TidyTuesday challenges | On [GitHub](https://github.com/rfordatascience/tidytuesday), [Twitter](https://twitter.com/search?q=%23TidyTuesday&src=typeahead_click).
- gganimate: [thomasp85/gganimate](https://github.com/thomasp85/gganimate)
- tidyexplain: [gadenbuie/tidyexplain](https://github.com/gadenbuie/tidyexplain)
- Distill (theme for RMarkdown): https://rstudio.github.io/distill/
- Google & [StackOverflow](https://stackoverflow.com/) are your best friends!
